---
title: Introduction to R
subtitle: Intensive Statistics Course
author:
  name: Wihan Marais
  affiliation: \text{wihanmarais@sun.ac.za} | github.com/WihanZA
date: "25 January 2022" ## Or "Lecture no."
output: 
  # html_document:
  #   keep_md: false
  #   theme: journal
  #   highlight: haddock
  #   # code_folding: show
  #   toc: yes
  #   toc_depth: 4
  #   toc_float: yes
  #   css: css/preamble.css ## For multi-col environments
  pdf_document:
    latex_engine: xelatex
    keep_tex: false ## Change to true if want keep intermediate .tex file
    toc: false
    toc_depth: 3
    dev: cairo_pdf
    # fig_width: 7 ## Optional: Set default PDF figure width
    # fig_height: 6 ## Optional: Set default PDF figure height
    extra_dependencies: ['booktabs', 'threeparttable', 'float'] # "longtable"
    includes:
      in_header: tex/preamble.tex ## For multi-col environments
    pandoc_args:
        --template=tex/mytemplate.tex ## For affiliation field. See: https://bit.ly/2T191uZ
always_allow_html: true
header-includes:
- \usepackage{siunitx}
- \usepackage{graphicx}
- \usepackage{float}
- \usepackage{cleveref}

urlcolor: blue
# mainfont: Cochineal
# sansfont: Fira Sans
# monofont: Fira Code
## Although, see: https://tex.stackexchange.com/q/294362
## Automatically knit to both formats:
knit: (function(inputFile, encoding) {
 rmarkdown::render(inputFile, encoding = encoding, 
 output_format = 'all') 
 })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, dpi=300, message = FALSE, warning = FALSE)
```

\graphicspath{ {./images/} }

## Why R?

\begin{enumerate}
\item R is free and open-source.
\begin{itemize}
\item At the time of writing, a new \href{https://www.quantec.co.za/store/}{Stata 17 annual license} is priced between R11,670 and R21,280 excluding VAT.
\item Free upgrades, updates and dissemination.
\item Availability of helpful resources like \textbf{\href{https://stackoverflow.com/}{stackoverflow}}. \end{itemize}
\item R uses \textbf{packages}
\begin{itemize}
\item R consists of Base-R coupled with third-party libraries of pre-written code, or packages.
\item \textbf{CRAN}, or The Comprehensive R Archive Network, is a network of ftp (file transfer protocol) and web servers around the world that store identical, up-to-date, versions of code and documentation for R.
\item More on this later.
\end{itemize}
\item R uses predictive coding (Ctrl/Cmd + Space is very useful).


\includegraphics[width=\textwidth]{pred_text}
\item R is compatible with Markdown.
\begin{itemize}
\item These lecture notes were created as a `.Rmd` file using \textbf{R Markdown}, RStudio's native authoring framework for data science.
\item See this \href{https://rmarkdown.rstudio.com/lesson-1.html}{1-minute video summary} of what R Markdown entails.
\end{itemize}
\end{enumerate}

## Before we start

You need the following installed on your machine:
\begin{itemize}
\item \textbf{\href{https://cran.r-project.org/}{R}} or Base-R.

"R is a freely available language and environment for statistical computing and graphics which provides a wide variety of statistical and graphical techniques: linear and nonlinear modelling, statistical tests, time series analysis, classification, clustering, etc."

\item \textbf{\href{https://www.rstudio.com/}{RStudio}}

"RStudio is an integrated development environment (IDE) for R. It includes a console, syntax-highlighting editor that supports direct code execution, as well as tools for plotting, history, debugging and workspace management."

\item \textbf{\href{https://cran.r-project.org/bin/windows/Rtools/rtools40.html}{Rtools}}

Select the Rtools download link for the relevant version of R installed on your machine. To determine the version currently installed, run the following code in your console. First, highlight the line of code you would like to run and Ctrl/Cmd + Enter to run.
\end{itemize}

```{r rversion, eval=FALSE}
sessionInfo()[1]$R.version$version.string

# IMPORTANT:
# Take care to check the box to have the installer 'edit your path'
```

To verify that we have installed Rtools properly, we need to make use of the \verb|devtools| package.
```{r devtools, eval = F}
install.packages("devtools") # Install the package from CRAN.
library(devtools) # Load package into your current library.

find_rtools() # Run this command from the devtools package
# or
devtools::find_rtools() 
# should return TRUE in your console
```
## Important packages

Install and load a few packages that you would likely use often. Let's use the \verb|installr| package as an example.

```{r pac_example1, eval = F}
# Packages need only be installed once
install.packages("installr")
# and can be loaded into your library with
library(installr)
```

However, if I have already installed \verb|installr| before, \verb|install.packages()| produces the following error:
\begin{figure}[H]
\includegraphics{install_pac}
\centering
\end{figure}

It can be hard to keep track of all the packages that you have or have not installed on your machine. How, then, should we install packages?

```{r pac_example2, eval = F}
# In short: if package has not yet been installed, run code to install
if (!require(installr)) {
    install.packages("installr")
    require(installr)
}
```

Instead, I propose using the \verb|pacman| package. It enables us to easily install new- and load old packages from curated lists such as CRAN, or any open-source package from \textbf{\href{https://github.com/}{GitHub}} using \verb|p_load()| and \verb|p_load_gh()|, respectively. These commands install packages if they have not yet been installed, and subsequently load them into our library. 

```{r pacman, eval = F}
# Installs pacman from CRAN.
if (!require(pacman)) {
    install.packages("pacman")
    require(pacman)
}
# Load pacman into our library.
library(pacman)

# And finally...
pacman::p_load(installr)
```

Why did we load \verb|installr| in the first place?

```{r installr, eval = F}
# Are you using the latest version of R?
check.for.updates.R()  
# Download and run the latest R Version.
install.R()
# Copy your packages to the newest R installation
copy.packages.between.libraries()
```

What are the packages we need to install and load?
```{r packages_setup, eval= F}
# From CRAN
pacman::p_load(fixest, tidyverse, huxtable, hrbrthemes, modelsummary, glue)

# From GitHub
pacman::p_load_gh("karthik/wesanderson",
                  "BlakeRMills/MetBrewer")
                # "profile/repository name"
```









### Regression tables

I have fairly strong preferences about how regression tables should look (threeparttable FTW). Luckily, the fantastic **modelsummary** package has us covered for nice looking regression tables, particularly since it automatically supports different Rmd output formats and backends. (For example, via the equally excellent **kableExtra** package.) This makes it easy to produce regression tables that look good in both HTML and PDF... although the latter requires that the corresponding LaTeX packages be loaded first. This template loads those LaTeX packages automatically, so tables like the below Just Work^TM^.

```{r msummary, message=FALSE, warning=FALSE}
library(fixest) ## For quick multi-model regression object

mods = feols(c(mpg, hp) ~ disp + csw(wt, drat) | cyl + vs, data = mtcars)

library(modelsummary)
library(kableExtra)

msummary(
  mods, 
  title = "fixest: multi-model estimation", 
  stars = TRUE,
  gof_omit = "Adj|Pseudo|Log|AIC|BIC"
  ) %>%
  add_footnote(
    c(paste("This footnote is pretty long. In fact, it runs over several lines",
          "of standard PDF output. Luckily that's no problem thanks to",
          "modelsummary, kableExtra, and threeparttable. As an aside, the",
          "fixest package is also amazing and you should use it."),
      "A shorter note."),
    threeparttable = TRUE
    ) %>%
  kable_styling(latex_options = "hold_position") ## (Optional) Print table directly below code
```

### PDF support for non-standard fonts

This is an easy one; simply a matter of adding `dev: cairo_pdf` to the YAML. But it's nice not having to remember that every time, no?

*Note: As the figure caption suggests, to run this next chunk you'll need to add [Arial Narrow](https://docs.microsoft.com/en-us/typography/font-list/arial-narrow) to your font book if it's not installed on your system already.*

```{r mpg, message=FALSE}
library(ggplot2)
library(hrbrthemes)

ggplot(mtcars, aes(mpg, wt)) +
  geom_point() +
  labs(x = "Fuel efficiency (mpg)", y = "Weight (tons)",
       title = "This plot uses Arial Narrow fonts",
       caption = "Note: Fonts must be installed separately on your system.") + 
  theme_ipsum()
```

## Acknowledgements and further reading

Lecture notes are compiled from the following resources:
\begin{itemize}
\item \textit{\href{https://raw.githack.com/grantmcdermott/R-intro/master/rIntro.html}{R intro}} (2018) by Grant R. McDermott and Ed Rubin.
\item \textit{\href{https://datsci.nfkatzke.com/posts/2019-07-24-practical-1-getting-you-started/}{Data Science for Economics and Finance: Getting you staRted}} (2021) by N.F. Katzke.
\end{itemize}
Should you need additional resources to get started, try the following:
\begin{itemize}
\item \href{https://www.statmethods.net/}{Quick-R}
\item \href{https://stata2r.github.io/}{Stata2R}
\item \href{https://stat447.com/}{Data Science Programming Methods (STAT 447)} by Dirk Eddelbuettel (University of Illinois)
\item \href{https://www.rstudio.com/resources/cheatsheets/}{RStudio Cheatsheets}
\item \href{https://github.com/uo-ec607/lectures}{Data Science for Economists (EC 607)} by  Grant McDermott (University of Oregon)
\item Use your student credentials to sign up for a \href{https://education.github.com/pack}{GitHub Pro} account. 
\item Download \href{https://desktop.github.com/}{GitHub Desktop} for free and use version control for all your projects.
\end{itemize}
 

 



