---
title: "19478844"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Task3, echo=FALSE}
pacman::p_load(fixest, tidyverse, huxtable, ucra, glue)

data1 <- read.csv(file = "data/ts_data.csv", header = TRUE)

data1 <- data1 %>% mutate(Year = as.Date(Year, format = "%Y-%m-%d")) %>%
  dplyr::filter(Year >= as.Date("1985-01-01"), Year <= as.Date("2010-01-01"), 
                Country == "France") %>%
  arrange(Year)

pacman::p_load(urca)

test_vector <- data1 %>% pull(ln_consump_pc)
my_ADF1 <- ur.df(
  y = test_vector, 
  type = "trend", 
  lags = 5, 
  selectlags = "AIC" 
)

summary(my_ADF1)

```


## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.height = 3.5, fig.width = 6)

library(pacman)


pacman::p_load(fixest, tidyverse, huxtable, ucra, glue)

data1 <- read.csv(file = "data/ts_data.csv", header = TRUE)

data1 <- data1 %>% mutate(Year = as.Date(Year, format = "%Y-%m-%d"))

dplyr::filter(Year >= as.Date("1985-01-01"), Year <= as.Date("2010-01-01"), country == "France") %>%
arrange(Year)

```{r}
##Task 4 
data2 <- read.csv(file = "data/cs_data.csv", header = TRUE)

pacman::p_load(fixest)

cs_df <- read.csv("Data/cs_data.csv")
model1 <- feols(wage ~ x1, data = cs_df)
model2 <- feols(wage ~ x1 + x2, data = cs_df)
model3 <- feols(wage ~ x1 + x2 + i(treat), data = cs_df)

huxreg("Model1" = model1, "Model2"= model2, "Model3" = model3,
       statistics = c("N" = "nobs", "R-squared" = "r.squared"),
       stars = c(`*`= 0.1, `**` = 0.05, `***` = 0.01, `****` = 0.001),
       number_format = 2
       ) %>%
  set_font_size(9)
```


